XSGRB - The X-shooter GRB afterglow sample

### 2017-01-03 document/XSGRB_sample.tex
Selection Criteria section

\usepackage{url}
\usepackage{color,hyperref}
\definecolor{linkcolor}{rgb}{0,0,0.7}
\hypersetup{colorlinks=true,
	linkcolor=linkcolor, 
	citecolor=linkcolor,
	filecolor=linkcolor, 
	urlcolor=linkcolor}

@@

\usepackage{xcolor}
\newcommand\todo[1]{\textcolor{red}{(#1)}}

@@

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, etc.. This is, however,
only possible as long as there are satellites in orbit that rapidly and
accurately locate GRBs. The currently operating \textit{Swift} satellite,
launched in 2004 and still fully-functioning, allows for very efficient
follow-up observations of GRBs due to its unprecedented rate, speed and
precision of localisations.

There are several unanswered but fundamental questions that must be addressed in
order to exploit the full potential of GRBs as cosmological probes. More than
50\% of the \textit{Swift} bursts with measured redshift are at $z > 2$, and
5--10\% are expected to be above $z=5$ \citep{Salvaterra2007,
	Salvaterra2012,Jakobsson2012, Perley2016}. A high redshift completeness is
crucial for our understanding of the link between the number density of GRBs per
unit redshift and the global star-formation history of the Universe, as measured
by other means \citep[UV, FIR, sub-mm, see][]{Robertson2012}. The detection of
GRBs at $z > 6$ shows that GRBs have become competitive as a tool to identifying
galaxies at the highest redshifts and unsurpassed in providing detailed
abundance information via absorption line spectroscopy \citep{Tanvir2012,
	McGuire2016}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists from already, being it weather conditions, pointing
restrictions of the telescope or poorly localized bursts as reported by the
\textit{Swift}-telescope. To maximize the return of the follow-up campaign
we have chosen a few selection criteria that attempts to provide an unbiased
selection of bursts, while allowing for a high success-rate

\footnote{Note that in the P84 proposal the criteria have
been stated a bit differently, the visibility constraint being replaced by a
declination + Sun angle constraint. The below criteria are however those
defining the sample.}

\begin{enumerate}
	\item GRB triggered onboard by Swift.
	\item Galactic $A_V \lesssim 0.5$ mag.
	\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12.
	\item No bright closeby stars.

\end{enumerate}

We simultaneously want to minimize any biases against astrophysical conditions
while at the same time maximizing likelihood of observations. By restricting
the selection criteria to conditions local to Galaxy.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex} and plot the positions off all the bursts on
the night sky in Fig. \ref{fig:skymap}. The central zone of avoidance due to
high galactic extinction cutoff is visible with an apparent isotropic
distribution of the bursts away from the galactic center except in the upper
right corner, where Earth casts a shadow as seen from the telescope.

@@

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/p_values.pdf}}
	\caption{Relational graph shown the .}
	\label{fig:p_values}
\end{figure}

We therefore conclude that the sample presented here, to a high degree
represents the intrinsic properties of the GRBs in the full \textit{Swift} sample.

>>>>

\usepackage{url}
\usepackage{color,hyperref}
\definecolor{linkcolor}{rgb}{0,0.3,0.7}
\hypersetup{colorlinks=true,
	linkcolor=linkcolor, 
	citecolor=linkcolor,
	filecolor=linkcolor, 
	urlcolor=linkcolor}

@@

\usepackage{xcolor}
\newcommand\todo[1]{\textbf{(#1)}}

@@

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006a, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, etc.. This is, however,
only possible as long as there are satellites in orbit that rapidly and
accurately locate GRBs. The currently operating \textit{Swift} satellite,
launched in 2004 and still fully-functioning, allows for very efficient
follow-up observations of GRBs due to its unprecedented rate, speed and
precision of localisations.

There are several unanswered but fundamental questions that must be addressed in
order to exploit the full potential of GRBs as cosmological probes. More than
50\% of the \textit{Swift} bursts with measured redshift are at $z > 2$, and
5--10\% are expected to be above $z=5$ \citep{Salvaterra2007,
	Salvaterra2012,Jakobsson2012, Perley2016a}. A high redshift completeness is
crucial for our understanding of the link between the number density of GRBs per
unit redshift and the global star-formation history of the Universe, as measured
by other means \citep[UV, FIR, sub-mm, see][]{Robertson2012}. The detection of
GRBs at $z > 6$ shows that GRBs have become competitive as a tool to identifying
galaxies at the highest redshifts and unsurpassed in providing detailed
abundance information via absorption line spectroscopy \citep{Tanvir2012,
	McGuire2016}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and follow-up}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists already, being it weather conditions, pointing restrictions
of the telescope or poorly localized bursts as reported by the alerting
telescope. To maximize the return of the follow-up campaign we have chosen a few
selection criteria that attempts to provide an unbiased selection of bursts,
while allowing for a high follow-up success rate. The importance of defining
unbiasing selection criteria has been highlighted previously
\citep{Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a}, when trying to
address the specific underlying distribution functions such as the redshift
distribution, host metallicity distribution, or brightness distribution. When
investigating a specific distribution function, a high degree of
\textit{completeness} is desired meaning that in terms of the parameter of
interest, the largest possible fraction of targets in the sample has
determinations of that parameter \citep{Perley2016b}.

We simultaneously want to minimize any biases against astrophysical conditions
while at the same time maximizing likelihood of observations allowing us to
obtain a higher degree of observing completeness. By restricting the selection
criteria to conditions local to Galaxy and therefore irrespective of intrinsic
GRB properties, the hope is that the sample collect represents the underlying
distribution of GRBs in a fair way while at the same time, additional unbiasing
cuts can be imposed that can improve the follow-up success. The selection
criteria used here is based on previous successes \cite{Jakobsson2006b,
	Fynbo2009, Hjorth2012}, which has shown to represent the underlying distribution
to a high degree. We test the degree of completeness in Sect.
\ref{completeness}.

The selection criteria are defined as follows:

\begin{enumerate}
	\item GRB triggered onboard by Swift.
	\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No bright closeby stars.
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

Because our ability to observe GRB afterglows is strongly correlated with the
rapidness and precision of the target positions delivered by the triggering
facilities, by selecting only bursts that have been triggered on board the
\textit{Swift} space telescope \cite{Gehrels2004}, by the Burst Alert Telescope
(BAT) \citep{Barthelmy2005}, we start out with a homogeneously selected sample
where burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006}, attempts at inferring properties of the
underlying cosmic GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016} and thereby by
restricting the follow-up effort to bursts detected by \textit{Swift}, the
limitations of the parent sample is well studied.

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, we additionally require an X-ray position from the X-ray
Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} with 10 hours and to account for observing constraints on
\textit{Swift} that XRT began observations within 10 minutes. The additional
rapidness requirement of the XRT follow-up means that the number of detected
X-ray afterglows for BAT triggers is 100 percent \citep{Burrows2007}. The
\textit{Swift} constraints on the GRB follow-up means that the sample presented
here inherits the \textit{Swift} sample properties, except the additional XRT
constraint. Because the XRT completeness is very high\citep{Burrows2007}, this
should change the parent sample significantly. This issue is addressed again in
Sect. \ref{completeness}.

To allow for observability we additionally require that the GRB is visible
(duh!) from the telescope site at Cerro Paranal, Antofagasta, Región de
Antofagasta, Chile, for a least 1 hour allowing time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
seen from Earth and that GRB properties does not depend on position on the sky
\citep{Briggs1996}, this cut does not influence our ability to sample the
underlying GRB population. The same arguments applies to the requirement that
there are no nearby bright stars, which is additionally supplemented by the
requirement what the galactic dust content does not exert an extinction higher
than $A_V \lesssim 0.5$ mag based on the extinction maps by
\citet{Schlegel1998}. Because of  isotropy, these additional cuts should not
influence the optical properties of the bursts themselves, only our ability to
successfully secure observations that allows us to investigate the spectroscopic
properties of GRBs.

We address the succesfullness of these criteria and the actual follow-up effort
in representing the full \textit{Swift} sample ion Sect. \ref{completeness}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure and practicalities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Will be written by Daniele?}

@@

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex} and plot the positions off all the bursts on
the night sky in Fig. \ref{fig:skymap}. The central zone of avoidance due to
high galactic extinction cutoff is visible with an apparent isotropic
distribution of the bursts away from the galactic center except in the upper
right corner, where Earth casts a shadow on the sky as seen from the telescope.

@@

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/XSGRB_p_values.pdf}}
	\caption{Relational graph shown the respective p-values that the different
	sample are drawn from the same underlying distribution. The arrows represent
	the comparison with each of the p-values are listed next to. Only in the HI
	column density distribution is there mild evidence against the null, but the
	discrepancy is mainly driven by a relatively larger fractional contribution
	from low-column hosts.}
	\label{fig:p_values}
\end{figure}

We therefore conclude that the sample presented here, to a high degree
represents the intrinsic properties of the GRBs in the full \textit{Swift} sample.
\todo{This doesn't quite feel exhaustive? Is there a better way of addressing the issue of completeness}
